Re-examining cross-cultural similarity judgments using lexical co-occurrence

### 2023-02-02 code/4_paper/manuscript/manuscript.tex
manuscript updates

\usepackage[utf8]{inputenc}
\usepackage[T1, T2A, T5]{fontenc}
\usepackage{kotex}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

An influential perspective within cultural psychology links these
differences in similarity judgment to tendencies toward analytic
processing in Western cultures, and holistic processing in East Asian
cultures (Nisbett, 2003). Analytic processing emphasizes rule-like
relationships predicated on objects and their properties and,
correspondingly, taxonomic similarity, while holistic processing
emphasizes relations between objects and their context and therefor,
thematic similarity. In related work, East Asian participants show a
higher level of sensitivity to context than their Western counterparts
when reproducing drawings from memory (Ji, Peng, \& Nisbett, 2000);
visually exploring naturalistic scenes (Chua, Boland, \& Nisbett, 2005);
describing scenes (Masuda \& Nisbett, 2001); and in explaining the
causes of ambiguous behaviors (Choi, Nisbett, \& Norenzayan, 1999). The
consensus interpretation in this literature ascribes cross-cultural
differences in similarity judgment to variation in the
\emph{conceptualization} of similarity -- with people from East Asian
cultures relying on a more thematic notion of similarity than
Westerners.

@@

We note that language statistics may incorporate both cultural-specific
environmental statistics, that is, the experiential \emph{input} to
similarity judgments, and culture-specific senses of similarity, the
\emph{conceptualization} of similarity. For example, cultural features
(like farming) can lead to differences in environmental statistics
(seeing cows and grass) and these can influence language (talking more
about cows and grass). But other cultural features (like conceiving of
similarity thematically) could also cause individuals to talk
differently about the same experiences (mentioning what cows eat rather
than what other animals cows are like). Acoordingly, our approach
examines the extent to which language statistics can predict
cross-cultural differences in similarity judgments with the
understanding that language statistics are likely a proxy for both
\emph{input} to and \emph{conceptualization} of similarity.

@@

The present study tests whether the statistics of language can be used
to predict cross-cultural differences in similarity judgments, and
particularly, whether this approach provides additional insight beyond
cross-cultural characterizations based on taxonomic vs thematic
preferences.

We measured taxonomic versus thematic similarity matching in a
forced-choice word triad task in three populations. Following Ji et al.
(2004), we measured preferences in the US and China. In addition, we
collected data in a novel context: Vietnam. Vietnam is a Southeast Asian
country that borders China and has historically been greatly influenced
by Chinese culture (Hui, 2002). Therefore, it serves as a suitable
cultural context to investigate whether the claim made by Ji et al.
(2004) and previous studies -- that Eastern and Western cultures have
different notions of similarity -- extends beyond mainland China. In
addition to these replication and extension questions, we tested whether
fastText vectors from corpora corresponding to each language context
(English, Mandarin, and Vietnamese) are good predictors for similarity
judgments in each population.

@@

We recruited 200 participants from the US, 199 participants from
Vietnam, and 200 participants from mainland China. US participants were
recruited through snowball sampling seeded with Stanford student email
lists, Vietnam participants were recruited through snowball sampling
seeded with Vietnam-based student groups on Facebook, and mainland China
participants were recruited through snowball sampling seeded with group
chats on WeChat. US participants were compensated with \$5 gift
certificates (USD), VN participants received 50,000₫ (VND) in phone
credit, and mainland China participants received 25CNY through WeChat
credit transfer.

We excluded 8 US participants, 62 Vietnam participants, and 16 China
participants who missed 2 or more attention checks. We followed 4
exclusion criteria that aim to retain only participants who are
influenced by one culture: (1) non-native speakers of English and
Vietnamese, respectively, (2) fluent in at least one of the other two
study languages (Vietnamese for US participants, English for Vietnamese
participants and Chinese participants), (3) have lived outside of the
test country (US, Vietnam, or China) for more than two years, and (4)
have significant international experience (more than 6 international
experiences of 2 days or longer.) We did not use a particular criterion
for a language if it would exclude 25\% or more of any one sample. In
this round of exclusion, we excluded 73 US, 27 Vietnam participants, and
35 China participants. After these exclusions, the US sample included
119 participants (30M, 84F, 3 non-binary, 2 other), with mean age = 22.2
(SD = 8.15) and median age = 20. The Vietnam sample included 110
participants (34M, 71F, 5 other), with mean age = 22.21 (SD = 5.81) and
median age = 21. The China sample included 149 participants (61M, 87F, 1
other), with mean age = 23.1 (SD = 3.65) and median age = 23.\footnote{A
  table summarizing number of participants lost at each round of
  exclusions is included in the Supplementary Information.}

We preregistered a more stringent exclusion criterion where participants
were excluded if they missed any attention checks. However, this led to
a small sample size, especially for Vietnam context (US = 109, China =
132, Vietnam = 57). Our reported results with the less stringent
exclusion criterion (detailed above) is largely not different from the
results with the preregistered criterion. Any differences are noted in
the Analysis section.

\hypertarget{stimuli-materials}{%
\subsection{Stimuli Materials}\label{stimuli-materials}}

We adapted stimuli from previous studies to create a set of test triads
consisting of a cue, with one thematic and one taxonomic match option.
For example, ``cow,'' ``grass,'' and ``chicken,'' where ``cow'' is the
cue, ``grass'' is the thematic match, and ``chicken'' the taxonomic
match. We included 105 such triads, a superset including triads pulled
from supplemental information and in-text examples across the
literature, and others that we adapted or created . We selected triads
on the basis of cultural familiarity in the US, Vietnam, and China. The
triads were originally in English; they were translated to Vietnamese
and Mandarin by a fluent bilingual speaker in each language. The
translations were then checked for accuracy after backtranslation to
English by another fluent bilingual in each language who was naive to
the original English versions. All materials are available at {[}BLINDED
FOR REVIEW{]}.

@@

Each participant completed all 105 triads in sets of 21 trials at a time
(10 test triads, 10 filler triads, and 1 attention check per page), by
selecting the match most related to the cue (``Which thing is most
closely related to the bolded item?'' and translated equivalents). The
test triads were presented with 105 filler triads mixed in, to obscure
the taxonomic-thematic two-answer forced choice structure of the test
stimuli and reduce the likelihood that participants would become aware
of the design. The filler triads were groups of three semantically
related words, but where the match options were not distinguished by
thematic vs.~taxonomic similarity, e.g., cue ``bird'' with match options
``lizard'' and ``toad.'' Additionally, we included 10 attention check
trials, which were formatted like the test and filler triads but
included an instruction instead of a cue item, e.g., ``Choose wife''
with match options ``wife'' and ``husband.'' In total, each participant
completed 210 similarity judgments and 10 attention check questions,
with triads presented in randomized orders that varied between subjects.

@@

To give an intuition for our model, consider again the cow-grass-chicken
triad: we retrieved word vectors for ``cow'' and ``grass'', and
calculate the cosine distance between these vectors. Similarly, we
retrieved vectors for ``cow'' and ``chicken'' and calculate the cosine
distance between them. Our similarity prediction is then inversely
proportional to the ratio of cosine distance of these pairs. This is
because a larger cosine distance means the word vectors are further
apart, and thus the words are less similar. For example, if the cosine
distance of thematic cow-grass is 0.7 and the cosine distance of
taxonomic cow-chicken is 0.3, then our model predicts, correspondingly,
that 30\% of responses to the triad will be grass, and the other 70\%
chicken.

In practice, we calculated the cosine distance between each cue-thematic
match (thematic cosine distance) and cue-taxonomic match (taxonomic
cosine distance), using the spatial.distance.cosine function from the
SciPy package (Virtanen et al., 2020). We then calculated the thematic
cosine distance proportion as thematic cosine distance over the sum of
taxonomic cosine distance and thematic cosine distance. We did this for
all three corpora. We were able to obtain predictions for all triads in
all languages. We then use a mixed-effects regression to evaluate how
well each corpus model predicts participants' similarity judgments,
across triads and cultural contexts. All analyses are available at
{[}BLINDED FOR REVIEW{]}.

@@

Overall, there is a significant effect of country on proportion of
thematic responses (\(\chi^2\)(2) = 15.37, p \textless{} .001). However,
this effect is driven by the difference between US and China responding
(\(\beta\) = -0.48, p \textless{} .001). There is no statistical
difference between the Vietnam and China responding (\(\beta\) = -0.22,
p = 0.09), and the US and Vietnam responding (\(\beta\) = 0.26, p =
0.086).

@@

To investigate whether language statistics predicts triads and filler
items differently, we compare the variance in responding that can be
accounted for by language statistics of the corresponding corpus in the
two different types of items (structured triads versus non-structured
filler items). Namely, we ran two mixed-effects logistic regression
models that predict responding across cultural contexts (1=thematic or
word1, 0=taxonomic or word2), with the corresponding corpus statistics
(for e.g., cosine proportion from the English corpus for US responding)
as a fixed effect and participant and item as random effects. Consistent
with our results above, corresponding corpus is a significant predictor
for responding to both triads and filler items (triads: \(\beta\) =
-1.79, \(\chi^2\)(1) = 75.58, p \textless{} .001, fillers: \(\beta\) =
-2, \(\chi^2\)(1) = 70.89, p \textless{} .001). Importantly, we found
comparable conditional \(R^2\) when corresponding corpus is used to
predict only triad items (\(R^2\) = 0.3) and only filler items (\(R^2\)
= 0.26), suggesting that culture-corresponding corpus statistics is able
to explain the same amount of variance in the responding data for either
triads or filler items. This result provides evidence for our view that
language statistics is measuring general tendencies in similarity
judgment, beyond preference for thematic/taxonomic relations.

@@

In this paper, we consider whether statistics of the environment (as
proxied by language statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. We replicated the previously documented
contrast between English speakers in the US and Mandarin Chinese
speakers from East Asia (mainland China, Taiwan, Hong Kong, and
Singapore), with mainland China participants preferring thematic
relations to a greater extent compared to US participants. Our sample of
Vietnamese participants showed intermediate response but not
significantly different from either Chinese or US participants. This
finding suggests some limitations on the generality of the cultural
account, which proposes that thematic/taxonomic similarity preference
aligns with East Asian/Western tendency for holistic vs analytic
processing. We found some support for the environmental statistics
account: each corpus statistics is a good predictor for the
corresponding country's similarity judgments, even when other corpus
statistics are included, and even with triads without a
thematic/taxonomic structure. Overall, our results provide evidence that
cross-cultural differences in similarity judgment are related to
linguistic co-occurrence patterns (which also vary across cultures).

@@

Ji et al. (2004) established that culture-aligned differences in this
paradigm exist, even when the test language is held constant, concluding
that ``it is culture (independent of the testing language) that led to
different grouping styles'' in their study. Our data provide a
cautionary note to this conclusion, suggesting that while cultural
differences in similarity judgment exists, we might get better traction
at modelling cross-cultural variation through environmental statistics
such as lexical co-occurrences, rather than treating groups of cultures
as monoliths (East Asian vs Western). However, there are still many open
questions for this account. Our operationalization of environmental
statistics with lexical co-occurrence is a proxy of both language inputs
and culture-specific conceptualization of similarity. The extent to
which cross-cultural differences in similarity judgment are driven by
language versus culture is still an open question. Future work should
also aim to provide a more specific computational account of how lexical
co-occurrence might guide categorization preference beyond the simple
proportion-of-similarity model tested here, and investigate other
individual- or triad-specific factors that might influence how
environmental statistics affect similarity judgment.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1, T2A, T5]{fontenc}
\usepackage{kotex}
\usepackage{CJKutf8}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

An influential perspective within cultural psychology links these
differences in similarity judgment to tendencies toward analytic
processing in Western cultures, and holistic processing in East Asian
cultures (Nisbett, 2003). Analytic processing emphasizes rule-like
relationships predicated on objects and their properties and,
correspondingly, taxonomic similarity, while holistic processing
emphasizes relations between objects and their context and therefore,
thematic similarity. In related work, East Asian participants show a
higher level of sensitivity to context than their Western counterparts
when reproducing drawings from memory (Ji, Peng, \& Nisbett, 2000);
visually exploring naturalistic scenes (Chua, Boland, \& Nisbett, 2005);
describing scenes (Masuda \& Nisbett, 2001); and in explaining the
causes of ambiguous behaviors (Choi, Nisbett, \& Norenzayan, 1999). The
consensus interpretation in this literature ascribes cross-cultural
differences in similarity judgment to variation in the
\emph{conceptualization} of similarity -- with people from East Asian
cultures relying on a more thematic notion of similarity than
Westerners.

@@

We note that language statistics may incorporate both cultural-specific
environmental statistics, that is, the experiential \emph{input} to
similarity judgments, and culture-specific senses of similarity, the
\emph{conceptualization} of similarity. For example, cultural features
(like farming) can lead to differences in environmental statistics
(seeing cows and grass) and these can influence language (talking more
about cows and grass). But other cultural features (like conceiving of
similarity thematically) could also cause individuals to talk
differently about the same experiences (mentioning what cows eat rather
than what other animals cows are like). Accordingly, our approach
examines the extent to which language statistics can predict
cross-cultural differences in similarity judgments with the
understanding that language statistics are likely a proxy for both
\emph{input} to and \emph{conceptualization} of similarity.

@@

The present study tests whether the statistics of language can be used
to predict cross-cultural differences in similarity judgments, and
particularly, whether this approach provides additional insight beyond
cross-cultural characterizations based on taxonomic vs.~thematic
preferences.

We measured taxonomic vs.~thematic similarity matching in a
forced-choice word triad task in three populations. Following Ji et al.
(2004), we measured preferences in the US and China. In addition, we
collected data in a novel context: Vietnam. Vietnam is a Southeast Asian
country that borders China and has historically been greatly influenced
by Chinese culture (Hui, 2002). Therefore, it serves as a suitable
cultural context to investigate whether the claim made by Ji et al.
(2004) and previous studies -- that Eastern and Western cultures have
different notions of similarity -- extends beyond mainland China. In
addition to these replication and extension questions, we tested whether
fastText vectors from corpora corresponding to each language context
(English, Mandarin, and Vietnamese) are good predictors for similarity
judgments in each population.

@@

Data collection and analyses for this study were pre-registered, and the
pre-registration is available at {[}blinded for peer review{]}. We
recruited 200 participants from the US, 200 from mainland China, and 199
from Vietnam. All participants were recruited through snowball sampling,
which, in the US was seeded with student email lists at a large
university, in China (CN) with student groups on WeChat, and in Vietnam
(VN) with Vietnam-based student groups on Facebook. US participants were
compensated with \$5 gift certificates (USD), CN participants received
25CNY through WeChat credit transfer, and VN participants received
50,000₫ (VND) in phone credit,

We excluded 8 US participants, 16 CN participants, and 62 VN
participants, who missed two or more of the 11 attention check
questions.

In doing so, we deviated from our preregistered exclusion criterion,
which used a more stringent policy, excluding participants who missed
any of the 10 attention checks. This approach would have substantially
reduced our sample size, disproportionately affecting the VN sample
(resulting in a final sample of 109 US, 132 CN, and 57 VN participants).
We report results with the less stringent exclusion criterion, but we
performed analyses with both samples and found broadly similar results.
Any differences are noted in the Results \& Discussion section.

We also applied four demographic exclusion criteria designed to reduce
cross-cultural influences across our samples: we excluded participants
who (1) were not native speakers of the test language (English,
Vietnamese, or Mandarin Chinese), (2) were fluent in another one of the
study languages, (3) had lived outside of the test country for more than
two years, or (4) had significant international experience (more than 6
international experiences of 2 days or longer). Following our
pre-registration, we dropped any exclusion criterion that would exclude
25\% or more of any one population, but only for that population.
Specifically, we did not exclude VN and CN participants who spoke
English, or US participants with significant international experience.
As a result of these demographic exclusions, we excluded 73 US, 35 CN,
and 27 VN participants. The final US sample included 119 participants
(30M, 84F, 3 non-binary, 2 other), with mean age = 22.2 (SD = 8.15). The
CN sample included 149 participants (61M, 87F, 1 other), with mean age =
23.1 (SD = 3.65). And the VN sample included 110 participants (34M, 71F,
5 other), with mean age = 22.21 (SD = 5.81).\footnote{A table
  summarizing the number of participants omitted for each criterion is
  included in the Supplementary Information.}

\hypertarget{materials}{%
\subsection{Materials}\label{materials}}

All stimuli and scripts for the experiment and analysis are available at
{[}BLINDED FOR REVIEW{]}. We collected and adapted stimuli from previous
studies to create test triads consisting of a cue with one thematic and
one taxonomic match option. For example, ``cow,'' ``grass,'' and
``chicken,'' where ``cow'' is the cue, ``grass'' is the thematic match,
and ``chicken'' the taxonomic match. We included 105 such triads, a
superset including triads pulled from supplemental information and
in-text examples across the literature, and others that we adapted or
created . We selected triads on the basis of cultural familiarity within
the US, China, and Vietnam. The triads were translated from English to
Mandarin and Vietnamese by fluent bilingual speakers of each language,
and backtranslated to English by another fluent bilingual who was naive
to the original English version. The backtranslations were checked
against the original English to identify ambiguities in the translation,
which were resolved through discussion and selection of alternative
terms when relevant.

@@

\begin{CJK*}{UTF8}{gbsn}
Each participant completed all 105 triads in blocks of 21 trials at a time (10 test triads, 10 filler triads, and 1 attention check per page), by selecting the match most related to the cue (following Ji et al., 2004): "Which thing is most closely related to the bolded item?" / "哪个事物与加粗项关联最紧密?" / "Thứ nào liên quan nhất với thứ được in đậm?".
The test triads were presented with 105 filler triads mixed in, to obscure the taxonomic-thematic two-answer forced choice structure of the test stimuli and reduce the likelihood that participants would become aware of the design.
The filler triads were groups of three semantically related words, but where the match options were not distinguished by thematic vs. taxonomic similarity, for example, the cue "bird" with match options "lizard" and "toad." Additionally, we included 11 attention check trials, which were formatted similarly to the test and filler triads but included an instruction instead of a cue item, e.g., "Choose wife" with match options "wife" and "husband." In total, each participant completed 210 similarity judgments (105 test triads and 105 fillers) and 11 attention check questions, with all items presented in randomized orders that varied between subjects.
\end{CJK*}

@@

To give an intuition for our model, consider again the cow:
grass/chicken triad. We retrieved word vectors for ``cow'' and
``grass'', and calculated the cosine distance between these vectors. We
did the same for ``cow'' and ``chicken.'' Our similarity prediction is
inversely proportional to the ratio of cosine distance of these pairs.
This is because a larger cosine distance means the word vectors are
further apart, and thus the words are less similar. For example, if the
cosine distance of thematic cow-grass is 0.7 and the cosine distance of
taxonomic cow-chicken is 0.3 (the more similar of the two), then our
model predicts, correspondingly, that 30\% of responses to the triad
will be grass, and the other 70\% chicken.

We calculated the cosine distance between each cue-thematic match
(thematic cosine distance) and cue-taxonomic match (taxonomic cosine
distance), using the spatial.distance.cosine function from the SciPy
package in R (Virtanen et al., 2020). We then calculated the thematic
cosine distance proportion as thematic cosine distance over the sum of
taxonomic cosine distance and thematic cosine distance. We followed this
process for all three corpora, and were able to obtain predictions for
all triads in all languages. Then, we used a mixed-effects regression to
evaluate how well each corpus model predicts participants' similarity
judgments, across triads and cultural contexts.

@@

Overall, there is a significant effect of country on proportion of
thematic responses (\(\chi^2\)(2) = 15.37, p \textless{} .001). However,
this effect is driven by the difference between US and China responding
(\(\beta_{US}\) = -0.48, p \textless{} .001). There is no statistical
difference between the Vietnam and China responding (\(\beta_{Vietnam}\)
= -0.22, p = 0.09), and the US and Vietnam responding (\(\beta_{US}\) =
0.26, p = 0.086).

@@

To investigate whether language statistics predicts triads and filler
items differently, we compare the variance in responding that can be
accounted for by language statistics of the corresponding corpus in the
two different types of items (structured triads vs.~non-structured
filler items). Namely, we ran two mixed-effects logistic regression
models that predict responding across cultural contexts (1=thematic or
word1, 0=taxonomic or word2), with the corresponding corpus statistics
(for e.g., cosine proportion from the English corpus for US responding)
as a fixed effect and participant and item as random effects. Consistent
with our results above, corresponding corpus is a significant predictor
for responding to both triads and filler items (triads: \(\beta\) =
-1.79, \(\chi^2\)(1) = 75.58, p \textless{} .001, fillers: \(\beta\) =
-2, \(\chi^2\)(1) = 70.89, p \textless{} .001). Importantly, we found
comparable conditional \(R^2\) when corresponding corpus is used to
predict only triad items (\(R^2\) = 0.3) and only filler items (\(R^2\)
= 0.26), suggesting that culture-corresponding corpus statistics is able
to explain the same amount of variance in the responding data for either
triads or filler items. This result provides evidence for our view that
language statistics is measuring general tendencies in similarity
judgment, beyond preference for thematic/taxonomic relations.

@@

In this paper, we consider whether statistics of the environment (as
proxied by language statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. We replicated the previously documented
contrast between English speakers in the US and Mandarin Chinese
speakers from East Asia (mainland China, Taiwan, Hong Kong, and
Singapore), with mainland China participants preferring thematic
relations to a greater extent compared to US participants. Our sample of
Vietnamese participants showed intermediate response but not
significantly different from either Chinese or US participants. This
finding suggests some limitations on the generality of the cultural
account, which proposes that thematic/taxonomic similarity preference
aligns with East Asian/Western tendency for holistic vs.~analytic
processing. We found some support for the environmental statistics
account: each corpus statistics is a good predictor for the
corresponding country's similarity judgments, even when other corpus
statistics are included, and even with triads without a
thematic/taxonomic structure. Overall, our results provide evidence that
cross-cultural differences in similarity judgment are related to
linguistic co-occurrence patterns (which also vary across cultures).

@@

Ji et al. (2004) established that culture-aligned differences in this
paradigm exist, even when the test language is held constant, concluding
that ``it is culture (independent of the testing language) that led to
different grouping styles'' in their study. Our data provide a
cautionary note to this conclusion, suggesting that while cultural
differences in similarity judgment exists, we might get better traction
at modelling cross-cultural variation through environmental statistics
such as lexical co-occurrences, rather than treating groups of cultures
as monoliths (East Asian vs.~Western). However, there are still many
open questions for this account. Our operationalization of environmental
statistics with lexical co-occurrence is a proxy of both language inputs
and culture-specific conceptualization of similarity. The extent to
which cross-cultural differences in similarity judgment are driven by
language vs.~culture is still an open question. Future work should also
aim to provide a more specific computational account of how lexical
co-occurrence might guide categorization preference beyond the simple
proportion-of-similarity model tested here, and investigate other
individual- or triad-specific factors that might influence how
environmental statistics affect similarity judgment.
